Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-01-24 ms.Rnw
much more in talk

\begin{thm}\label{thm:pruned_lik}
  Let $\P^*_T=(y^*,z^*)$  be a given pruned genealogy and
  \begin{equation}\label{eq:phidef}
    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z'),
  \end{equation}
  where $n$ is the deme occupancy (\cref{sec:demes}), $\ell$ and $s$ are the lineage-count and saturation functions, respectively (\cref{sec:ells}), $Q$ is the compatibility indicator (\cref{sec:compatibility}), and the binomial ratio is as defined in \cref{sec:binomial_ratio}.
  Then
  \begin{equation*}
    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,y^*_t,\zt^*_t,z^*_t)}.
  \end{equation*}
\end{thm}
\begin{proof}
  If $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\Hr_T$ and $\P^*_T$ are incompatible and $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*_T$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(y^*_t,z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t,\zt^*_t,z^*_t)$ are emergent.
  By assumption, at each genealogical event, lineages within a deme are exchangeable:
  each has an identical probability of being involved.
  This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $\P^*_t$.
  Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
  \begin{equation*}
    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(y^*_t,z^*_t)}{r^{\U_t}}{s(\yt^*_t,y^*_t,\zt^*_t,z^*_t)}}.
  \end{equation*}
  Returning to the possibility that $\H_T$ is incompatible with $\P^*_T$, since $\Prob{\Pr_T=\P^*_T}=0$ if either any $Q_u=0$ or any event of $\P^*_T$ is not an event of $\H_T$, we obtain the result.
\end{proof}

@@

\begin{thm}
  Suppose that $\P^*_T=(y^*,z^*)$ is a pruned genealogy, for $T>0$.
  If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)=
      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}
      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,y^*_t,\zt^*_t,z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  where $\phi$ is defined in \cref{eq:phidef},
  then the likelihood of $\P^*_T$ is
  \begin{equation*}
    \lik(\P^*_T)=\int{w(T,x)\,\dd{x}}.
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:obsc_lik}
  Let $z^*=\obs(\P^*_T)$ be a given obscured genealogy.
  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{z^*}}{q(x,y)}=1$.
  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,z^*_t)=1$
  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,y^{\Pr_T}_0=y}>0$.
  Then there is a Markov jump process $\yr_t$ on $\Yspace_{z^*}$ such that
  \begin{equation*}
    \CondProb{\zr_T=z^*}{\Hr_T=\H_T}=
    \Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
  \end{equation*}
  where $\phi$ is defined in \cref{eq:phidef} and
  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
  \end{equation}
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
  \begin{equation*}
    \begin{gathered}
      \Prob{\yrh_0}=q(\yrh_0),
      \qquad
      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
    \end{gathered}
  \end{equation*}
  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.

@@

\begin{thm}
  Let $z^*$, be a given obscured genealogy.
  Let $\pi$, $q$, and $\yr_t$ be as in \cref{thm:obsc_lik}.
  For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,z^*_t)}{\pi_u(t,x,x',y,y')},\\[2ex]
      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),
    \end{gathered}
  \end{equation*}
  where $\phi$ is defined in \cref{eq:phidef}.
  Suppose $w=w(t,x,y)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x,y)=
      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}\\
      &-\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
      &+\sum_{e\in\event{\z}{}^*}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y,y')\,\psi_u(t,x',x,y',y)\,\dd{x'}}}-w(t,x,y)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
  Then the likelihood of $V^*_T$ is
  \begin{equation*}
    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
  \end{equation*}
\end{thm}

>>>>

\begin{thm}\label{thm:pruned_lik}
  Let $\P^*_T=(\y^*,\z^*)$  be a given pruned genealogy and
  \begin{equation}\label{eq:phidef}
    \phi_u(x,y,y',z,z')\coloneq\BinRatio{n(x)}{\ell(y',z')}{r^{u}}{s(y,y',z,z')}\,Q_{u}(y,y',z,z'),
  \end{equation}
  where $n$ is the deme occupancy (\cref{sec:demes}), $\ell$ and $s$ are the lineage-count and saturation functions, respectively (\cref{sec:ells}), $Q$ is the compatibility indicator (\cref{sec:compatibility}), and the binomial ratio is as defined in \cref{sec:binomial_ratio}.
  Then
  \begin{equation*}
    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\Indicator{\event{\H}{T}\supseteq{\event{\P}{T}^*}}\,\prod_{t\in\event{\H}{T}}{\phi_{\U_t}(\X_t,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}.
  \end{equation*}
\end{thm}
\begin{proof}
  If $\event{\Hr}{T}\nsupseteq\event{\P}{T}^*$, then $\Hr_T$ and $\P^*_T$ are incompatible and $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
  Similarly, if any event of $\Hr_T$ is incompatible with the local structure of $\P^*_T$ in the sense of \cref{sec:compatibility}, then $\CondProb{\Pr_T=\P^*_T}{\Hr_T}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Conditional on $\Hr_T=\H_T$, at each time $t\in\event{\H}{T}$, a jump of mark $\U_t$ occured, with a production of $r^{\U_t}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(\X_t)=(n_i)_{i\in\Demes}$.
  In $\P^*_T$, at time $t$, there are $\ell_i=\ell_i(\y^*_t,\z^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)$ are emergent.
  By assumption, at each genealogical event, lineages within a deme are exchangeable:
  each has an identical probability of being involved.
  This exchangeability implies that each lineage present in a deme at time $t$ was equally likely to have been one of the emergent lineages.
  In particular, at time $t$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ red balls and $n_i-r_i$ black balls, exactly $s_i$ of the drawn balls are red, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $\P^*_t$.
  Moreover, since, again conditional on $\Hr_T$, the identities of the lineages involved in a genealogical event are random and independent of the identities selected at all other events, we have established that
  \begin{equation*}
    \CondProb{\Pr_T=\P^*_T}{\Hr_T=\H_T}=\prod_{t\in\event{\H}{T}}{\BinRatio{n(\X_t)}{\ell(\y^*_t,\z^*_t)}{r^{\U_t}}{s(\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)}}.
  \end{equation*}
  Returning to the possibility that $\H_T$ is incompatible with $\P^*_T$, since $\Prob{\Pr_T=\P^*_T}=0$ if either any $Q_u=0$ or any event of $\P^*_T$ is not an event of $\H_T$, we obtain the result.
\end{proof}

@@

\begin{thm}
  Suppose that $\P^*_T=(y^*,z^*)$ is a pruned genealogy, for $T>0$.
  If $w=w(t,x)$ satisfies the initial condition $w(0,x)=p_0(x)$ and the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)=
      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}}
      -\sum_u{\int{w(t,x)\,\alpha_u(t,x,x')\,\dd{x'}}}\\
      &\quad+\sum_{e\in\event{\P}{T}^*}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(x,\yt^*_t,\y^*_t,\zt^*_t,\z^*_t)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  where $\phi$ is defined in \cref{eq:phidef},
  then the likelihood of $\P^*_T$ is
  \begin{equation*}
    \lik(\P^*_T)=\int{w(T,x)\,\dd{x}}.
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:obsc_lik}
  Let $\z^*=\obs(\P^*_T)$ be a given obscured genealogy.
  Let $\pi$ and $q$ be probability kernels, such that for all $u\in\Jumps$, $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $\pi_u(t,x,x',y,y')\ge{0}$, $\sum_{y'\in\Yspace_t^{z^*}}{\pi_u(t,x,x',y,y')}=1$, $q(x,y)\ge{0}$, and $\sum_{y\in\Yspace_0^{\z^*}}{q(x,y)}=1$.
  Suppose moreover that $\pi_u(t,x,x',y,y')>0$ whenever $Q_u(y,y',\zt^*_t,\z^*_t)=1$
  and that $q(x,y)>0$ whenever $\Prob{\Xr_0=x,\y^{\Pr_T}_0=y}>0$.
  Then there is a Markov jump process $\yr_t$ on $\Yspace_{\z^*}$ such that
  \begin{equation*}
    \CondProb{\zr_T=\z^*}{\Hr_T=\H_T}=
    \Indicator{\event{\H}{T}\supseteq\event{\z}{T}^*}\,\Expect{\frac{1}{q(X_0,\yr_0)}\,\prod_{t\in\event{\H}{T}}{\frac{\phi_{\U_t}(\X_t,\yrt_t,\yr_t,\zt^*_t,\z^*_t)}{\pi_{\U_t}(t,\Xt_t,\X_t,\yrt_t,\yr_t)}}},
  \end{equation*}
  where $\phi$ is defined in \cref{eq:phidef} and
  the expectation is taken over the sample paths of $\yr_t$, $t\in[0,T]$.
\end{thm}
\begin{proof}
  First, observe that, since $\obs$ is a deterministic operator,
  \begin{equation}\label{eq:IS1}
    \CondProb{\zr_T=z^*}{\Hr_T}=\CondExpect{\Indicator{\obs(\Pr_T)=z^*}}{\Hr_T}.
  \end{equation}
  Our strategy will be to evaluate \cref{eq:IS1} using importance sampling:
  we will propose pruned genealogies compatible with $z^*_t$ as sample paths from a Markov process on $\Yspace_{z^*}$ and
  evaluate the the expectation in \cref{eq:IS1} by summing over these paths.
  Conditional on $\Hr_T$, the initial distribution $q$ and probability kernel $\pi$ generate a Markov chain, $\yrh_k$, on $\Yspace_{z^*}$ such that
  \begin{equation*}
    \begin{gathered}
      \Prob{\yrh_0}=q(\yrh_0),
      \qquad
      \CondProb{\yrh_k}{\yrh_{k-1},\Hr_T=\H_T}=\pi_{\Uh_k}(\Th_k,\Xh_{k-1},\Xh_{k},\yrh_{k-1},\yrh_k).
    \end{gathered}
  \end{equation*}
  The required Markov process $\yr_t$ is the unique \cadlag\ process with event times $\Trh_k$ and $\yrh_k$ as its embedded chain.
  This construction of $\yr_t$ obviously guarantees that $\event{\H}{t}\supseteq\event{\y}{}\supseteq\event{\z}{}^*$.

@@

\begin{thm}
  Let $z^*$, be a given obscured genealogy.
  Let $\pi$, $q$, and $\yr_t$ be as in \cref{thm:obsc_lik}.
  For $t\in\Rp$, $x,x'\in\Xspace$, $y,y'\in\Yspace^{z^*}_t$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y'),\\[2ex]
      \psi_u(t,x,x',y,y')=\frac{\phi_u(x',y,y',\zt^*_t,\z^*_t)}{\pi_u(t,x,x',y,y')},
    \end{gathered}
  \end{equation*}
  where $\phi$ is defined in \cref{eq:phidef}.
  Suppose $w=w(t,x,y)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x,y)=
      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\psi_u(t,x',x,y',y)\,\dd{x'}}}\\
      &-\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
      &+\sum_{e\in\event{\z}{}^*}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y,y')\,\psi_u(t,x',x,y',y)\,\dd{x'}}}-w(t,x,y)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x,y)=p_0(x)\,q(x,y)$.
  Then the likelihood of $\z^*_T$ is
  \begin{equation*}
    \lik(\z^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
  \end{equation*}
\end{thm}
